### 2024-07-13 base-rates.qmd
20240713 - more context

There are various ways to account for [base rates](#sec-baseRate).
One approach is to use [actuarial formulas](#sec-actuarialPrediction) (rather than human judgment) to make the predictions.
[Actuarial formulas](#sec-actuarialPrediction) based on [multiple regression](#sec-multipleRegression) or [machine learning](#sec-machineLearning) can account for the [base rate](#sec-baseRate) of the event.

Another approach is to leverage Bayes' theorem, using Bayesian updating and the probability nomogram.

@@

Bayesian updating involves comparing the relative probability of two outcomes, $P(C | R)$ versus $P(\text{not } C | R)$.
If we want to compare the relative probability of two outcomes, we can use the odds form of Bayes' theorem, as in @eq-bayes5:
$$
\begin{aligned}
  P(C | R) &= \frac{P(R | C) \cdot P(C_i)}{P(R_i)} \\
  P(\text{not } C | R) &= \frac{P(R | \text{not } C) \cdot P(\text{not } C_i)}{P(R_i)} \\
  \frac{P(C | R)}{P(\text{not } C | R)} &= \frac{\frac{P(R | C) \cdot P(C_i)}{P(R_i)}}{\frac{P(R | \text{not } C) \cdot P(\text{not } C_i)}{P(R_i)}} \\
  &= \frac{P(R | C) \cdot P(C_i)}{P(R | \text{not } C) \cdot P(\text{not } C_i)} \\
  &= \frac{P(C_i)}{P(\text{not } C_i)} \times \frac{P(R | C)}{P(R | \text{not } C)} \\
  \text{posterior odds} &= \text{prior odds} \times \text{likelihood ratio}
\end{aligned}
$$ {#eq-bayes5}

>>>>

There are various ways to account for [base rates](#sec-baseRate), including the use of [actuarial formulas](#sec-accountForBaseRatesActuarial) and the use of [Bayesian updating](#sec-bayesianUpdating).

### Actuarial Formula {#sec-accountForBaseRatesActuarial}

One approach to account for [base rates](#sec-baseRate) is to use [actuarial formulas](#sec-actuarialPrediction) (rather than [human judgment](#sec-humanJudgment)) to make the predictions.
[Actuarial formulas](#sec-actuarialPrediction) based on [multiple regression](#sec-multipleRegression) or [machine learning](#sec-machineLearning) can account for the [base rate](#sec-baseRate) of the event.

@@

Another approach to account for [base rates](#sec-baseRate) is to leverage Bayes' theorem, using Bayesian updating and the [probability nomogram](#sec-probabilityNomogram).
Bayesian updating is a form of [anchoring and adjustment](#sec-heuristicsAnchoringAdjustment); however, unlike the [anchoring and adjustment heuristic](#sec-heuristicsAnchoringAdjustment), it is a systematic approach to [anchoring and adjustment](#sec-heuristicsAnchoringAdjustment) that anchors one's predictions to the base rate, and then adjusts according to new information.
That is, we start with a [pretest probability](#sec-baseRate) (i.e., [base rate](#sec-baseRate)) and update our predictions based on the extent of new information (i.e., the [likelihood ratio](#sec-diagnosticLikelihoodRatio)).

To perform Bayesian updating involves comparing the relative probability of two outcomes, $P(C | R)$ versus $P(\text{not } C | R)$.
If we want to compare the relative probability of two outcomes, we can use the odds form of Bayes' theorem, as in @eq-bayes5:
$$
\begin{aligned}
  P(C | R) &= \frac{P(R | C) \cdot P(C_i)}{P(R_i)} \\
  P(\text{not } C | R) &= \frac{P(R | \text{not } C) \cdot P(\text{not } C_i)}{P(R_i)} \\
  \frac{P(C | R)}{P(\text{not } C | R)} &= \frac{\frac{P(R | C) \cdot P(C_i)}{P(R_i)}}{\frac{P(R | \text{not } C) \cdot P(\text{not } C_i)}{P(R_i)}} \\
  &= \frac{P(R | C) \cdot P(C_i)}{P(R | \text{not } C) \cdot P(\text{not } C_i)} \\
  &= \frac{P(C_i)}{P(\text{not } C_i)} \times \frac{P(R | C)}{P(R | \text{not } C)} \\
  \text{posterior odds} &= \text{prior odds} \times \text{likelihood ratio}
\end{aligned}
$$ {#eq-bayes5}
